---
title: "Testing a Sentinel-1 SAR preprocessing pipeline with spatialtestR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Testing a Sentinel-1 SAR preprocessing pipeline with spatialtestR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## The problem

Imagine you're building a **snow-on-runoff (SOR)** detection pipeline for the Alps from Sentinel-1 SAR data. The raw VH backscatter, in decibels, goes through five preprocessing steps before you can extract the wet-snow signal:

1.  **mask NoData**: convert sensor-encoded `-9999` to proper NA
2.  **remove bottom-noise**: discard pixels below `-35 dB`
3.  **speckle filter**: smooth the multiplicative noise typical of SAR
4.  **aggregate**: go from `10 m` to `20 m` resolution to reduce noise further
5.  **detect wet snow**: flag pixels whose backscatter drops more than `2 dB` below a winter reference

Each of these steps *should* preserve specific properties of the raster, and each can silently introduce bugs that look fine on a map but are numerically incorrect. This vignette walks through testing each step with `spatialtestR`.

```{r}
library(spatialtestR)
library(terra)
```

## Simplified SAR processing functions

> For this tutorial we use simplified versions to illustrate the test patterns.

```{r}
mask_nodata <- function(r, nodata = -9999) {
  v <- terra::values(r)
  v[v == nodata] <- NA
  terra::values(r) <- v
  r
}

mask_bottom_noise <- function(r, threshold_db = -35) {
  v <- terra::values(r)
  v[v < threshold_db] <- NA
  terra::values(r) <- v
  r
}

speckle_filter <- function(r, window = 3) {
  terra::focal(r,
               w = matrix(1 / (window * window), window, window),
               fun = sum, na.rm = TRUE,
               pad = TRUE, padValue = NA)
}

aggregate_raster <- function(r, factor = 2, fun = "mean") {
  terra::aggregate(r, fact = factor, fun = fun, na.rm = TRUE)
}

detect_wet_snow <- function(r, reference, threshold_db = 2.0) {
  reference - r > threshold_db
}
```

## Step 1: `mask_nodata()`

> convert the sentinel value `-9999` to proper `NA`. This should change only the `NA` pattern of the raster, nothing else.

```{r}
# Build a synthetic SAR-like raster: 200 NoData cells, 200 valid cells (dB)
r_in <- create_test_raster(
  nrow = 20, ncol = 20,
  crs    = "EPSG:32632",
  values = c(rep(-9999, 200), rep(-15, 200))
)

r_out <- mask_nodata(r_in, nodata = -9999)
```

> we expect the function to preserve geometry and CRS, and to introduce exactly 200 new `NA` values:

```{r}
# Structure: shape, CRS, layer count
expect_same_dims(r_in, r_out)
expect_same_crs(r_in,  r_out)

# Robustness: no Inf values, exactly 200 new NAs
expect_no_inf(r_out)
stopifnot(sum(is.na(terra::values(r_out))) == 200L)
```

> if we wanted to verify exactly what the function did to the values, we could ask `summarize_transformation()` for the regression-based view:

```{r}
summarize_transformation(r_in, r_out)
```

You'll see that R² is reported on the valid pixel pairs (excluding the masked NAs) and the slope/intercept identify this as a near-identity transformation **⟶** the function preserves values, it just relocates some of them to `NA`{style="color: rgb(76, 79, 105);"}.

## Step 2: `mask_bottom_noise()`

> below `-35 dB`, SAR returns are dominated by sensor noise rather than ground signal. This function masks those pixels to `NA`.

```{r}
# Mix of valid pixels (around -15 dB), some below -35 dB, some NA
set.seed(1)
r_in <- create_test_raster(
  nrow = 20, ncol = 20,
  crs    = "EPSG:32632",
  values = c(rep(-15,  200),   # valid
             rep(-40,  100),   # bottom-noise: will be masked
             rep(NA,   100))   # already NA
)

r_out <- mask_bottom_noise(r_in, threshold_db = -35)
```

> after this step we expect:
>
> - geometry unchanged
> - no values below the threshold
> - valid range now within typical SAR backscatter bounds
> - no Inf or NaN

```{r}
expect_same_dims(r_in, r_out)
expect_same_crs (r_in, r_out)

# After bottom-noise removal, all remaining values must be > -35 dB
expect_raster_values_between(r_out, lower = -35, upper = 5)

# No infinities introduced
expect_no_inf(r_out)
```

> confirming that the function adds `NAs` but leaves values otherwise untouched:

```{r}
summarize_transformation(r_in, r_out)
```

> in the output you should see roughly **100 added NAs** (the `-40 dB` pixels) and a linear fit classification of `identity` on the valid pairs.

## Step 3: `speckle_filter()`

> Speckle is the multiplicative noise pattern characteristic of SAR data. A typical filter (here: a simple 3×3 mean) smooths it out. **Crucially**, this is the first step in our pipeline where R² should drop below 1.

```{r}
# Build a noisy backscatter raster: mean -15 dB, sd 3 dB
set.seed(42)
r_in <- create_test_raster(
  nrow = 30, ncol = 30,
  crs    = "EPSG:32632",
  values = rnorm(900, mean = -15, sd = 3)
)

r_out <- speckle_filter(r_in, window = 3)
```

> what should be true after the filter:
>
> - **geometry preserved** (focal operations don't resize)
> - **no Inf** introduced
> - the **mean** stays close to the input mean (the filter is locally averaging, so it should be unbiased over a large enough region)

```{r}
expect_same_dims(r_in, r_out)
expect_same_crs (r_in, r_out)
expect_no_inf(r_out)

# Mean should be preserved within a small tolerance
expect_mean_preserved(r_in, r_out, tolerance = 0.5)
```

> `summarize_transformation()` will tell us whether the filter behaves as a "linear-ish" operation = slope close to 1, intercept close to 0, but R² *below* 1 because of the smoothing:

```{r}
summarize_transformation(r_in, r_out)
```

**A typical result:** `R² ≈ 0.4–0.6` (lots of smoothing), classification `non_linear` (because we cross below the `linear_threshold = 0.999`). That is *exactly* what you want from a working filter.

> **Reading R² for a filter**: R² close to 1 would mean the filter isn't actually filtering anything. R² close to 0 would mean the output is uncorrelated with the input (bug). Real speckle filters land in between.

## Step 4: `aggregate_raster()`

> now the geometry actually changes: we go from a 30×30 raster to a 15×15 raster by averaging 2×2 blocks. `expect_same_dims()` would (correctly) fail here, so this is the case for **`summarize_transformation()` with the `resampling` argument**.

```{r}
r_in  <- create_test_raster(nrow = 30, ncol = 30,
                            crs = "EPSG:32632",
                            values = rnorm(900, mean = -15, sd = 3),
                            seed = 7)
r_out <- aggregate_raster(r_in, factor = 2, fun = "mean")
```

> we can no longer compare pixel-to-pixel: the rasters have different dimensions.\
> `summarize_transformation()` switches to **geographic sampling** automatically:

```{r}
summarize_transformation(r_in, r_out,
                         resampling = "average",
                         seed = 1)
```

> the output includes a *"resampling hint"* section that tells you the expected R² range for the chosen method (here `"average"` → roughly 0.92 - 0.99) and whether your observed value falls within that range.

If we were to assert *no* resampling happened we'd get a hard error:

```{r, error = TRUE}
summarize_transformation(r_in, r_out, resampling = FALSE)
```

## Step 5: `detect_wet_snow()`

> comparing the preprocessed raster to a **winter reference** backscatter and flags pixels whose backscatter has dropped more than 2 dB (an established wet-snow indicator). The function returns a boolean raster.

```{r}
# Build a winter reference (slightly higher backscatter, mean -12 dB)
ref <- create_test_raster(
  nrow = 30, ncol = 30,
  crs    = "EPSG:32632",
  values = rnorm(900, mean = -12, sd = 2),
  seed   = 11
)

# Build a "spring" raster with some genuinely wet pixels
spring <- create_test_raster(
  nrow = 30, ncol = 30,
  crs    = "EPSG:32632",
  values = c(rnorm(450, mean = -12, sd = 2),   # not wet
             rnorm(450, mean = -16, sd = 2)),  # wet (below threshold)
  seed = 12
)

wet <- detect_wet_snow(spring, reference = ref, threshold_db = 2.0)
```

> now we test the wet-snow mask:

```{r}
# Structure: same shape and CRS as the input
expect_same_dims(spring, wet)
expect_same_crs (spring, wet)

# A boolean raster: all values must be 0 or 1
expect_raster_values_between(wet, lower = 0, upper = 1)

# No Inf or NaN values
expect_no_inf(wet)
```

> We can verify the relationship between input and output too: A non-trivial wet-snow detection should produce output that is *not* a linear function of input (it's a thresholded comparison, not a smooth transformation):

```{r}
summarize_transformation(spring, wet)
```

You should see a classification like `non_linear`: perfectly correct for a thresholded comparison. The diagnostic confirms what you already know structurally: this is a thresholding operation, not a value transformation.

## Error messages

One of the design goals of `spatialtestR` is that when an assertion fails, you get an error that **tells you exactly what went wrong**. Below are the typical failure modes you'll encounter in a SAR pipeline, each shown with the actual error output.

### A function that silently changes dimensions

Imagine `mask_nodata()` accidentally trimmed a row from the raster:

```{r, error = TRUE}
r_in  <- create_test_raster(nrow = 20, ncol = 20, crs = "EPSG:32632")
r_bad <- create_test_raster(nrow = 19, ncol = 20, crs = "EPSG:32632")  # one row lost

expect_same_dims(r_in, r_bad)
```

### A function that silently reprojects

`expect_same_crs()` catches accidental CRS changes:

```{r, error = TRUE}
r_in  <- create_test_raster(nrow = 10, ncol = 10, crs = "EPSG:32632")
r_bad <- create_test_raster(nrow = 10, ncol = 10, crs = "EPSG:4326")    # silently reprojected

expect_same_crs(r_in, r_bad)
```

### Values out of the expected range

If your `mask_bottom_noise()` left a few `-40 dB` pixels behind, this assertion catches it and reports the observed range:

```{r, error = TRUE}
r_buggy <- create_test_raster(
  nrow = 10, ncol = 10,
  values = c(rep(-15, 90), rep(-40, 10))   # 10 pixels below -35 dB
)

expect_raster_values_between(r_buggy, lower = -35, upper = 5)
```

### Inf values from a division-by-zero bug

If your speckle filter or normalisation pipeline divides by zero, you'll get `Inf` cells. `expect_no_inf()` reports the counts of `Inf` and `-Inf` separately:

```{r, error = TRUE}
r_buggy <- create_test_raster(nrow = 10, ncol = 10, values = 1:100)
v <- terra::values(r_buggy)
v[c(1, 2, 3)]  <- Inf
v[c(50)]        <- -Inf
terra::values(r_buggy) <- v

expect_no_inf(r_buggy)
```

### Systematic mean shift

A speckle filter should preserve the mean of an extended homogeneous region. If your implementation has a systematic bias, `expect_mean_preserved()` flags it and reports the magnitude of the shift:

```{r, error = TRUE}
r_in   <- create_test_raster(values = rep(-15, 100))
r_buggy <- create_test_raster(values = rep(-13, 100))   # +2 dB systematic shift

expect_mean_preserved(r_in, r_buggy, tolerance = 0.5)
```

### Multiple mismatches in one error

When several properties are wrong at once, `spatialtestR` reports **all of them in a single error message**:

```{r, error = TRUE}
r_in  <- create_test_raster(nrow = 20, ncol = 20)
r_bad <- create_test_raster(nrow = 10, ncol =  5)    # both wrong

expect_raster_dims(r_in, nrow = 20, ncol = 20)       # this passes
expect_raster_dims(r_bad, nrow = 20, ncol = 20)      # this catches both at once
```

### When you assert no resampling but the geometry differs

Asserting `resampling = FALSE` on rasters with different dimensions raises a clear, descriptive error that prevents silent geometry drift from going unnoticed:

```{r, error = TRUE}
r_in  <- create_test_raster(nrow = 30, ncol = 30, crs = "EPSG:32632",
                            values = rnorm(900, mean = -15, sd = 3), seed = 1)
r_out <- terra::aggregate(r_in, fact = 2, fun = "mean")

summarize_transformation(r_in, r_out, resampling = FALSE)
```

------------------------------------------------------------------------

## Putting it all together: a single test suite

Below is the full pipeline test:

```{r}
library(spatialtestR)
library(terra)

# Synthetic Sentinel-1 fixture
r_raw <- create_test_raster(
  nrow = 30, ncol = 30,
  crs    = "EPSG:32632",
  values = c(rep(-9999, 200),
             rnorm(700, mean = -15, sd = 3)),
  seed = 1
)

# Run the full pipeline
r_step1 <- mask_nodata(r_raw)
r_step2 <- mask_bottom_noise(r_step1, threshold_db = -35)
r_step3 <- speckle_filter(r_step2, window = 3)
r_step4 <- aggregate_raster(r_step3, factor = 2, fun = "mean")

# Test all four geometry-preserving steps at once
for (step in list(r_step1, r_step2, r_step3)) {
  expect_same_dims(r_raw, step)
  expect_same_crs (r_raw, step)
  expect_no_inf(step)
}

# Test step 4 separately: geometry has changed (15x15 from 30x30)
expect_raster_dims(r_step4, nrow = 15, ncol = 15)
expect_same_crs (r_raw, r_step4)
expect_no_inf(r_step4)
```